---
title: "Latent class analysis of survey data in R — a worked blueprint"
subtitle: "Recoding, measurement checks, class enumeration, robustness checks and figures"
author: "Julia Gorny"
date: "2026"
output:
  github_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
params:
  data_file: "data/survey_data.sav"
  latentgold_dir: "data/latentgold"
  k_final: 4
  nrep: 10
  run_robustness: true
  m_imputations: 5
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = TRUE,
                      fig.width = 9, fig.height = 5.5, dpi = 150)

library(poLCA)    # loaded first: it attaches MASS, whose select() would
                  # otherwise mask dplyr::select()
library(dplyr)
library(tidyr)
library(purrr)
library(ggplot2)
library(haven)

source(here::here("R", "functions.R"))
source(here::here("R", "simulate_data.R"))
set.seed(123)
```

## About this analysis

This is a complete, runnable template for a latent class analysis of
attitudinal survey data in R, developed in the course of applied survey
research and generalised here. The structure, the modelling decisions and the
robustness checks are the ones used in practice; indicators carry neutral
names (`attitude_*`, `policy_*`, `norms_*`) and the data is synthetic, so the
workflow can be read, run and adapted on its own terms.

It covers the full path from raw survey items to publication figures:

1. Recoding non-response, reversing negatively worded items, collapsing
   sparse response categories
2. Measurement checks — Cronbach's alpha, a one-factor CFA, category shares
3. Export of a clean indicator set for estimation in LatentGOLD
4. Class enumeration in `poLCA` across candidate indicator sets, comparing
   BIC, AIC, relative entropy and minimum class size
5. Robustness checks: survey weights, multiple imputation of missing values,
   and sensitivity to the indicator set, with classes matched across
   solutions via the Hungarian algorithm
6. Weighted figures: conditional response probabilities, class composition
   by a categorical covariate, socio-demographic profiles, age distributions

**Data.** No survey data is included. The notebook runs on a synthetic
dataset generated by `R/simulate_data.R`, which builds a known four-class
structure so that every method below has something real to recover. All
numbers and figures here are illustrative only. To use it on your own data,
place a `.sav` file at `params$data_file` with matching variable names.

```{r load-data}
use_real_data <- file.exists(here::here(params$data_file))

raw <- if (use_real_data) {
  read_sav(here::here(params$data_file))
} else {
  simulate_survey(n = 1200)
}

cat(if (use_real_data) "Using the survey data." else
      "Survey data not found - running on SYNTHETIC data.",
    "\nRespondents:", nrow(raw), "\n")
```

## 1. Sample description (weighted)

```{r sample}
raw <- raw |>
  mutate(age_group = cut(age, breaks = c(15, 24, 34, 44, 54, 64, 74, 100),
                         labels = c("15-24", "25-34", "35-44", "45-54",
                                    "55-64", "65-74", "75+")))

weighted_shares(raw, "sex")
weighted_shares(raw, "age_group")
weighted_shares(raw, "vote", exclude = c(98, 99, 997, 999))
weighted_shares(raw, "education_raw", exclude = c(98, 99, 997, 999))
```

## 2. Recoding and measurement

### 2.1 Recoding rules

All indicator-building steps are collected in one function so that the
identical rules can be applied to the observed data and to the imputed data
in the robustness checks. The decisions are:

* **Non-response.** "Don't know" (code 6 on 5-point items, 999 on the 0–10
  policy items) is treated as missing rather than as a substantive answer.
  Sensitivity to this choice is tested in Section 5.2.
* **Direction.** Negatively worded items are reversed so that high values
  always mean higher endorsement within each block. `reverse_scale()` refuses values outside the valid range,
  so an unrecoded non-response code can never be turned into a valid answer.
* **Summated scale.** `scale_mean` is the mean of three items measuring the
  same construct, justified by Cronbach's alpha and a one-factor CFA
  (Section 2.2).
* **Collapsing.** All indicators are collapsed to three categories
  (low / medium / high). The 5- and 11-point items have many sparsely
  populated categories (Section 2.3), which LCA cannot estimate reliably, and
  three categories keep a substantively meaningful neutral middle.

```{r recoding-rules}
dk6_items   <- c("attitude_14", "attitude_15", "policy_7", paste0("norms_", 1:7),
                 paste0("norms_", 9:11), "norms_8")
dk999_items <- paste0("policy_", 1:6)

fem_reverse   <- c("attitude_1", "attitude_5", "attitude_6", "attitude_8", "attitude_9",
                   "attitude_10", "attitude_11", "attitude_12", "attitude_13", "attitude_14")
demo_reverse  <- c(paste0("norms_", 1:7), "norms_8", "norms_11")
five_point_cat <- c("attitude_1_r", "attitude_2", "attitude_3", "attitude_4",
                    "attitude_5_r", "attitude_6_r", "attitude_7", "attitude_8_r",
                    "attitude_9_r", "attitude_10_r", "attitude_11_r", "attitude_12_r",
                    "attitude_13_r", "attitude_14_r", "attitude_15", "policy_7",
                    "norms_12", paste0("norms_", 1:7, "_r"), "norms_8_r",
                    "norms_9", "norms_10", "norms_11_r")

prepare_indicators <- function(df) {
  df |>
    # non-response -> NA
    mutate(across(all_of(dk6_items),   ~ recode_to_na(.x, 6)),
           across(all_of(dk999_items), ~ recode_to_na(.x, 999)),
           norms_12 = recode_to_na(norms_12, 6:999)) |>
    # summated attitude scale (1-7 items)
    mutate(scale_mean = rowMeans(across(c(scale_item_1, scale_item_2, scale_item_3),
                                            as.numeric), na.rm = TRUE),
           scale_mean = ifelse(is.nan(scale_mean), NA, scale_mean),
           scale_cat = collapse_3(scale_mean, 3, 5, 7)) |>
    # reverse negatively worded items
    mutate(across(all_of(c(fem_reverse, demo_reverse)),
                  ~ reverse_scale(.x, 1, 5), .names = "{.col}_r")) |>
    # collapse to low / medium / high
    mutate(across(all_of(five_point_cat), ~ collapse_3(.x, 2, 3, 5),
                  .names = "{.col}_cat"),
           across(all_of(dk999_items), ~ collapse_3(.x, 3, 6, 10),
                  .names = "{.col}_cat"))
}

prepare_covariates <- function(df) {
  df |>
    mutate(
      income_cat = case_when(
        income_raw %in% 1:4   ~ "Low",
        income_raw %in% 5:11  ~ "Middle",
        income_raw %in% 12:16 ~ "High"),
      education_cat = factor(case_when(
        education_raw %in% 1:4  ~ "No to lower secondary education",
        education_raw %in% 5:7  ~ "Upper secondary and vocational post-secondary",
        education_raw %in% 8:10 ~ "Tertiary education"),
        levels = c("No to lower secondary education",
                   "Upper secondary and vocational post-secondary",
                   "Tertiary education")),
      # binary covariate: category 3 vs all other valid answers;
      # only 'not asked / missing' codes are excluded
      binary_cov = case_when(
        vote == 3 ~ 1,
        vote %in% c(999, 9998, 9999) ~ NA_real_,
        TRUE ~ 0),
      sex_f = as_factor(sex))
}

dat <- raw |> prepare_indicators() |> prepare_covariates()
```

### 2.2 Summated scale

```{r scale-check}
reliability_check(dat, c("scale_item_1", "scale_item_2", "scale_item_3"))

cfa_fit <- lavaan::cfa("Scale =~ scale_item_1 + scale_item_2 + scale_item_3",
                       data = dat, estimator = "MLR", std.lv = TRUE)
lavaan::standardizedSolution(cfa_fit) |>
  filter(op == "=~") |>
  select(item = rhs, loading = est.std, se, pvalue)
```

With three indicators the one-factor model is just-identified, so the CFA
is used to inspect the standardised loadings rather than global fit.

### 2.3 Coherence of indicator blocks and category sparsity

```{r blocks}
blocks <- list(
  block_attitude = c("attitude_1_r", "attitude_2", "attitude_3", "attitude_4", "attitude_5_r",
               "attitude_6_r", "attitude_7", "attitude_8_r", "attitude_9_r",
               "attitude_11_r", "attitude_10_r"),
  block_policy = c(paste0("policy_", 1:6), "policy_7"),
  block_norms = c(paste0("norms_", 1:7, "_r"), "norms_8_r",
                          "norms_9", "norms_10", "norms_11_r")
)

imap_dfr(blocks, ~ reliability_check(dat, .x) |> mutate(block = .y),
         .id = NULL) |>
  relocate(block)
```

Any pair above roughly r = 0.7 is worth dropping one of: near-collinear
indicators violate the local independence assumption and tend to produce a
class that separates on nothing but that redundancy.

```{r sparsity}
map_dfr(blocks, ~ category_shares(dat, .x), .id = "block") |>
  print(n = Inf)
```

Categories below about 10% of the sample are hard for LCA to estimate: the
conditional probabilities become unstable and solutions start depending on
the random start. Where several items have such categories, collapsing to
three is usually the cleaner choice, and it keeps a meaningful middle.

## 3. Indicator set and export to LatentGOLD

The final indicator set combines the summated attitude scale, one further
attitude item, seven policy-support items and nine items on institutional
norms. Indicators are given readable names here because variables cannot be
renamed inside LatentGOLD.

```{r export}
final_indicators <- c(
  attitude_scale                       = "scale_cat",
  attitude_single_item                 = "attitude_6_r_cat",
  policy_support_7                     = "policy_7_cat",
  policy_support_1                   = "policy_1_cat",
  policy_support_2        = "policy_2_cat",
  policy_support_3            = "policy_3_cat",
  policy_support_4       = "policy_4_cat",
  policy_support_5              = "policy_5_cat",
  policy_support_6      = "policy_6_cat",
  norm_item_1   = "norms_1_r_cat",
  norm_item_2      = "norms_2_r_cat",
  norm_item_3   = "norms_3_r_cat",
  norm_item_4      = "norms_4_r_cat",
  norm_item_5              = "norms_5_r_cat",
  norm_item_6          = "norms_6_r_cat",
  norm_item_7                  = "norms_7_r_cat",
  norm_item_8                 = "norms_9_cat",
  norm_item_9                 = "norms_8_r_cat"
)

missing_summary <- dat |>
  summarise(respondents_with_any_missing_pct =
              100 * mean(!complete.cases(across(all_of(final_indicators)))),
            item_cells_missing_pct =
              100 * mean(is.na(as.matrix(across(all_of(final_indicators))))))
round(missing_summary, 1)

dir.create(here::here("output"), showWarnings = FALSE)
dat |>
  select(ID, all_of(final_indicators), sex, age, education_cat,
         vote, group_var, weight) |>
  write_sav(here::here("output", "LCA_latentgold_input.sav"))
```

Typically a substantial share of respondents have at least one missing
indicator even when the share of missing item responses is small — which is
exactly the case where listwise deletion is wasteful. All models below
therefore keep these respondents (full-information ML, `na.rm = FALSE`)
rather than dropping them listwise.

## 4. Exploratory model selection in R

Before estimating the final weighted model in LatentGOLD, candidate indicator
sets were compared in `poLCA` (unweighted). For each set, models with 2–6
classes are fitted with `r params$nrep` random starts. Class enumeration
considers BIC, relative entropy (Ramaswamy et al., 1993), the smallest class
share and substantive interpretability together, not BIC alone.

```{r candidate-sets}
norms_items <- c(paste0("norms_", 1:7, "_r_cat"), "norms_9_cat",
                     "norms_10_cat", "norms_11_r_cat", "norms_8_r_cat")

candidate_sets <- list(
  block_attitude      = c("scale_cat", "attitude_1_r_cat", "attitude_2_cat",
                          "attitude_3_cat", "attitude_4_cat", "attitude_6_r_cat",
                          "attitude_7_cat", "attitude_8_r_cat", "attitude_9_r_cat",
                          "attitude_10_r_cat"),
  set_policy_attitude  = c("scale_cat", "policy_7_cat",
                          paste0("policy_", 1:6, "_cat")),
  set_policy_attitude_extra = c("scale_cat", "policy_7_cat",
                               paste0("policy_", 1:6, "_cat"),
                               "attitude_15_cat", "attitude_13_r_cat"),
  democracy           = norms_items,
  set_norms_attitude  = c(norms_items, "scale_cat", "attitude_1_r_cat",
                          "attitude_2_cat", "attitude_6_r_cat", "attitude_8_r_cat",
                          "attitude_10_r_cat"),
  set_norms_policy = c("policy_7_cat", paste0("policy_", 1:6, "_cat"),
                          norms_items),
  set_norms_policy_extra = c("policy_7_cat", paste0("policy_", 1:6, "_cat"),
                                norms_items, "attitude_15_cat",
                                "attitude_13_r_cat"),
  final               = unname(final_indicators)
)

candidate_fits <- map(candidate_sets, ~ fit_lca_range(dat, .x, ks = 2:6,
                                                       nrep = params$nrep))

map_dfr(candidate_fits, fit_table, .id = "indicator_set") |>
  print(n = Inf)
```

### 4.1 Final indicator set

```{r final-model}
final_fit <- candidate_fits$final[[paste0("K", params$k_final)]]$model

tibble(class = seq_along(final_fit$P),
       share_pct = round(100 * final_fit$P, 1))

dat$class_polca <- final_fit$predclass

cond_probs <- tidy_cond_probs(final_fit) |>
  mutate(item = names(final_indicators)[match(item, final_indicators)])

plot_cond_probs(cond_probs)
```

```{r class-demographics}
dat |>
  group_by(class = class_polca) |>
  summarise(weighted_n   = round(sum(weight)),
            male_pct     = round(100 * weighted.mean(sex == 1, weight), 1),
            mean_age     = round(weighted.mean(age, weight), 1),
            tertiary_pct = round(100 * weighted.mean(
              education_cat == "Tertiary education", weight, na.rm = TRUE), 1),
            binary_cov_pct = round(100 * weighted.mean(binary_cov, weight,
                                                      na.rm = TRUE), 1))
```

### 4.2 Local independence

LCA assumes that indicators are independent within classes. As a heuristic
check, within-class polychoric correlations above 0.7 are listed (LatentGOLD's
bivariate residuals were used for the final model).

```{r local-independence, warning = FALSE}
check_local_independence(dat, unname(final_indicators), "class_polca")
```

## 5. Robustness checks

```{r robustness-switch, include = FALSE}
knitr::opts_chunk$set(eval = params$run_robustness)
```

### 5.1 Survey weights

`poLCA` has no weights argument. As a workaround, each respondent is
replicated in proportion to their weight (weights × 10, rounded, which keeps
weights such as 3.13 and 3.96 distinct). This yields pseudo-maximum-likelihood
point estimates (Patterson, Dayton & Graubard, 2002), but it inflates the
log-likelihood and N roughly tenfold, which would make BIC favour ever more
classes. Information criteria are therefore rescaled to the original sample
size.

```{r weights}
weight_multiplier <- 10
dat_rep <- dat |>
  mutate(rep_n = round(weight * weight_multiplier)) |>
  filter(rep_n > 0) |>
  uncount(rep_n)

weighted_fits <- fit_lca_range(
  dat_rep, unname(final_indicators),
  ks = (params$k_final - 1):(params$k_final + 1), nrep = params$nrep,
  llik_scale = nrow(dat) / nrow(dat_rep), n_eff = nrow(dat))

fit_table(weighted_fits)

weighted_model <- weighted_fits[[paste0("K", params$k_final)]]$model

# classify the original respondents with the weighted model and match its
# classes to the unweighted solution before comparing class sizes
class_weighted <- apply(
  poLCA.posterior(weighted_model, y = as.matrix(dat[unname(final_indicators)])),
  1, which.max)
weights_alignment <- align_classes(dat$class_polca, class_weighted)

weights_alignment$mapping |>
  mutate(unweighted_pct = round(100 * final_fit$P[as.integer(class_a)], 1),
         weighted_pct   = round(100 * weighted_model$P[as.integer(class_b)], 1))
cat("Agreement of class assignments:", weights_alignment$agreement_pct, "%\n")
```

### 5.2 Missing data: multiple imputation

To check whether treating "don't know" as missing drives the result, the raw
items feeding the final indicators are multiply imputed (proportional-odds
models for ordinal items), the indicators are rebuilt with the same rules,
and the final model is re-estimated on each imputed dataset. Agreement with
the main solution is measured after optimally matching class labels.

```{r imputation}
items_to_impute <- c("scale_item_1", "scale_item_2", "scale_item_3", "attitude_6", "policy_7",
                     dk999_items, paste0("norms_", 1:7), "norms_9",
                     "norms_8")

imp_input <- raw |>
  mutate(across(all_of(intersect(items_to_impute, dk6_items)),
                ~ recode_to_na(.x, 6)),
         across(all_of(dk999_items), ~ recode_to_na(.x, 999))) |>
  select(all_of(items_to_impute)) |>
  mutate(across(everything(), ~ factor(.x, ordered = TRUE)))

imp <- mice::mice(imp_input, m = params$m_imputations, method = "polr",
                  printFlag = FALSE, seed = 123)

imputed_agreement <- map_dfr(seq_len(params$m_imputations), function(i) {
  completed <- mice::complete(imp, i) |>
    mutate(across(everything(), ~ as.numeric(as.character(.x))))
  d <- raw
  d[items_to_impute] <- completed
  d <- prepare_indicators(d)
  m <- poLCA(lca_formula(unname(final_indicators)), data = d,
             nclass = params$k_final, nrep = params$nrep, maxiter = 5000,
             na.rm = FALSE, verbose = FALSE)
  tibble(imputation = i,
         agreement_with_main_pct =
           align_classes(dat$class_polca, m$predclass)$agreement_pct)
})
imputed_agreement
```

Treating "don't know" as the neutral middle category instead was tested in a
separate script and is not reproduced here.

### 5.3 Indicator set: adding norms_12

The final model was re-estimated with an additional pluralism item
(`norms_12`, cultural enrichment). If the LatentGOLD class predictions of both
models are available, they are compared directly; otherwise both models are
estimated in `poLCA`.

```{r add-4c1}
lg_without <- file.path(here::here(params$latentgold_dir), "class_predictions_base.sav")
lg_with    <- file.path(here::here(params$latentgold_dir), "class_predictions_extra_item.sav")

if (use_real_data && file.exists(lg_without) && file.exists(lg_with)) {
  class_without <- read_sav(lg_without)[["clu#"]]
  class_with    <- read_sav(lg_with)[["clu#"]]
} else {
  class_without <- dat$class_polca
  class_with <- poLCA(lca_formula(c(unname(final_indicators), "norms_12_cat")),
                      data = dat, nclass = params$k_final, nrep = params$nrep,
                      maxiter = 5000, na.rm = FALSE, verbose = FALSE)$predclass
}

alignment <- align_classes(class_without, class_with)
alignment$crosstab
alignment$mapping
cat("Agreement after matching class labels:", alignment$agreement_pct, "%\n")
```

```{r robustness-switch-off, include = FALSE}
knitr::opts_chunk$set(eval = TRUE)
```

## 6. Figures for the article

The article's figures are based on the weighted LatentGOLD solution. If its
output is available it is used here; otherwise the figures are drawn from the
`poLCA` solution above.

```{r figure-data}
lg_classes <- file.path(here::here(params$latentgold_dir),
                        "class_assignments_final.sav")
lg_probs   <- file.path(here::here(params$latentgold_dir), "probabilities_lca.csv")

if (use_real_data && file.exists(lg_classes)) {
  fig_dat <- read_sav(lg_classes) |>
    rename(class = `clu#`) |>
    mutate(education_cat = as_factor(education_cat),
           sex_f = as_factor(sex),
           binary_cov = case_when(
             vote == 3 ~ 1,
             vote %in% c(999, 9998, 9999) ~ NA_real_,
             TRUE ~ 0))
} else {
  fig_dat <- dat |> mutate(class = class_polca)
}

if (use_real_data && file.exists(lg_probs)) {
  fig_probs <- read.csv(lg_probs) |>
    pivot_longer(starts_with("Prob_"), names_to = "class",
                 values_to = "probability") |>
    transmute(item = Indicator,
              class = as.integer(gsub("\\D", "", class)),
              category = factor(c("Low", "Medium", "High")[Answer],
                                levels = c("Low", "Medium", "High")),
              probability)
} else {
  fig_probs <- cond_probs
}
```

```{r fig-probabilities, fig.height = 7}
plot_cond_probs(fig_probs)
```

```{r fig-covariate}
# a categorical covariate crossed with class membership; swap in your own
# categories and colours
group_labels  <- c("1" = "Group A", "2" = "Group B", "3" = "Group C",
                   "4" = "Group D", "5" = "Group E")
group_colours <- c("Group A" = "#4C6E91", "Group B" = "#A3473F",
                   "Group C" = "#5C8A64", "Group D" = "#8A6FA8",
                   "Group E" = "#C08A3E")

plot_group_shares(fig_dat, group_labels, group_colours)
```

```{r fig-profiles}
plot_class_profiles(fig_dat)
```

```{r fig-age}
plot_age_by_class(fig_dat)
```

## References

Linzer, D. A., & Lewis, J. B. (2011). poLCA: An R package for polytomous
variable latent class analysis. *Journal of Statistical Software*, 42(10), 1–29.

Patterson, B. H., Dayton, C. M., & Graubard, B. I. (2002). Latent class
analysis of complex sample survey data: Application to dietary data.
*Journal of the American Statistical Association*, 97(459), 721–741.

Ramaswamy, V., DeSarbo, W. S., Reibstein, D. J., & Robinson, W. T. (1993). An
empirical pooling approach for estimating marketing mix elasticities with
PIMS data. *Marketing Science*, 12(1), 103–124.

## Session info

```{r session-info}
sessionInfo()
```
